---
output: 
  github_document:
    fig_width: 4.5
    fig_heigh: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, eval=TRUE,  cache=TRUE)
library(Rirt)
```

# Rirt: data analysis and parameter estimation using item response theory 

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/Rirt)](https://cran.r-project.org/package=Rirt)
<!-- badges: end -->


## Overview

Rirt provides an interface for conducting data analysis and parameter estimation using item respone theory (IRT) efficiently in [R](https://www.r-project.org/). A variety of unidimensional IRT models is included, such as the 3-parameter-logstic (3PL) model, the generalized partial credit model (GPCM), the graded response model (GRM), and a mixture of the aforementioned models.


## Installation

Install the stable version from CRAN:
```{r, eval=FALSE}
install.packages("Rirt")
```

Install the most recent version from [github](https://github.com/xluo11/Rirt):
```{r, eval=FALSE}
devtools::install_github("xluo11/Rirt")
```


## Quickstart

A common interface is created to make the package easy to use:

+ `model_[3pl/gpcm/grm/mixed]_prob`: compute the response probability for the given ability and item parameters
+ `model_[3pl/gpcm/grm/mixed]_info`: compute the information for the given ability and item parameters
+ `model_[3pl/gpcm/grm/mixed]_lh`: compute the response likelihood for the given ability and item parameters
+ `model_[3pl/gpcm/grm/mixed]_gendata`: generate response data and parameters from an IRT model
+ `model_[3pl/gpcm/grm]_plot`: plot item characteristic curves (ICC) or item informaiton functions (IIFs)
+ `model_[3pl/gpcm/grm]_rescale`: rescale ability and item parameters to a given scale
+ `model_[3pl/gpcm/grm/mixed]_eap`: score response vectors using the EAP method
+ `model_[3pl/gpcm/grm/mixed]_map`: score response vectors using the MAP method
+ `model_[3pl/gpcm/grm]_jmle`: estimate parameters using the joint maximum likelihood estimation method
+ `model_[3pl/gpcm/grm]_mmle`: estimate parameters using the marginal maximum likelihood estimation method


For example, users can quickly generate data from an IRT model of choice:
```{r}
# generate 10 people and 5 items using the 3PL model
# returns a list of 'u' (responses), 't' (ability), 'a' (item discrimination), 'b' (item difficulty), 'c' (pseudo-guessing)
model_3pl_gendata(n_p=10, n_i=5)

# generate 10 people and 5 4-category items using the GPCM model
# returns a list of 'u' (responses), 't' (ability), 'a' (item discrimination), 'b' (item difficulty), 'd' (category difficulty)
model_gpcm_gendata(n_p=10, n_i=5, n_c=4)

# generate 10 people and 5 4-category items using the GRM model
# returns a list of 'u' (responses), 't' (ability), 'a' (item discrimination), 'b' (item difficulty)
model_grm_gendata(n_p=10, n_i=5, n_c=4)
```

Compute the response probability, information, and likelihood using the 3PL model:
```{r}
# generate data
data <- model_3pl_gendata(n_p=10, n_i=5)
# compute P, returns a 2d array: people, item
with(data, model_3pl_prob(t, a, b, c, D=1.702))
# compute I, returns a 2d array: people, item
with(data, model_3pl_info(t, a, b, c, D=1.702))
# compute L, returns a 2d array: people, item
with(data, model_3pl_lh(u, t, a, b, c, D=1.702, log=FALSE))
```

Compute the response probability, information, and likelihood using the GPCM model:
```{r}
# generate data
data <- model_gpcm_gendata(n_p=10, n_i=5, n_c=3)
# compute P, returns a 3d array: people, item, category
with(data, model_gpcm_prob(t, a, b, d))
# compute I, returns a 3d array: people, item, category
with(data, model_gpcm_info(t, a, b, d))
# compute L, returns a 2d array: people, item
with(data, model_gpcm_lh(u, t, a, b, d, D=1.702, log=FALSE))
```

Compute the response probability, information, and likelihood using the GRM model:
```{r}
# generate data
data <- model_grm_gendata(n_p=10, n_i=5, n_c=3)
# compute P, returns a 3d array: people, item, category
with(data, model_grm_prob(t, a, b))
# compute I, returns a 3d array: people, item, category
with(data, model_grm_info(t, a, b))
# compute L, returns a 2d array: people, item
with(data, model_grm_lh(u, t, a, b, D=1.702, log=FALSE))
```

Compute the response probability, information, and likelihood using a mixture of IRT models:
```{r}
# generate data, returns response, theta, and a list of items
data <- model_mixed_gendata(n_p=10, n_3pl=5, n_gpcm=3, n_grm=3, n_c=3)
# compute P, returns a list
with(data, model_mixed_prob(t, items))
# compute I, returns a list when combine=FALSE or a 2d array when combine=TRUE
with(data, model_mixed_info(t, items, combine=TRUE))
# compute L, returns a list when combine=FALSE or a 2d array when combine=TRUE
with(data, model_mixed_lh(u, t, items, combine=TRUE))
```

## Visualization

Visualization is essential in IRT data analysis, and this package makes it really easy to produce high-quality characteristic curves and information function curves at both item and test level. 

For example, users can draw ICCs and IIFs for a set of 3PL items:
```{r}
data <- model_3pl_gendata(1, 5)
# item characteristic curves
with(data, model_3pl_plot(a, b, c, type='prob', total=FALSE))
# test characteristic curves
with(data, model_3pl_plot(a, b, c, type='prob', total=TRUE))
# item information funciton curves
with(data, model_3pl_plot(a, b, c, type='info', total=FALSE))
# test information function curves
with(data, model_3pl_plot(a, b, c, type='info', total=TRUE))
```

Draw ICCs and IIFs for a set of GPCM items:
```{r, fig.width=9, fig.height=3}
# Reproduce Figure 1 in Muraki's 1992 APM article
b <- matrix(c(-2, 0, 2, -.5, 0, 2, -.5, 0, 2), nrow=3, byrow=TRUE)
model_gpcm_plot(a=c(1, 1, .7), b=rowMeans(b), d=rowMeans(b)-b, D=1.0, d0=0)
# Reproduce Figure 2 in Muraki's 1992 APM article
b <- matrix(c(.5, 0, NA, 0, 0, 0), nrow=2, byrow=TRUE)
model_gpcm_plot(a=.7, b=rowMeans(b, na.rm=TRUE), d=rowMeans(b, na.rm=TRUE)-b, D=1.0, d0=0)
# Reproduce Figure 3 in Muraki's 1992 APM article
b <- matrix(c(1.759, -1.643, 3.970, -2.764), nrow=2, byrow=TRUE)
model_gpcm_plot(a=c(.778,.946), b=rowMeans(b), d=rowMeans(b)-b, D=1.0, d0=0)
```

```{r, fig.width=9, fig.height=5}
# Reproduce Figure 1 in Muraki's 1993 APM article
b <- matrix(c(0, -2, 4, 0, -2, 2, 0, -2, 0, 0, -2, -2, 0, -2, -4), nrow=5, byrow=TRUE)
model_gpcm_plot(a=1, b=rowMeans(b), d=rowMeans(b)-b, D=1.0)
# Reproduce Figure 2 in Muraki's 1993 APM article
b <- matrix(c(0, -2, 4, 0, -2, 2, 0, -2, 0, 0, -2, -2, 0, -2, -4), nrow=5, byrow=TRUE)
model_gpcm_plot(a=1, b=rowMeans(b), d=rowMeans(b)-b, D=1.0, type='info', item_level=TRUE)
```

Draw ICCs and IIFs for a set of GRM items:
```{r, fig.width=9, fig.height=3}
data <- model_grm_gendata(1, 3, n_c=3)
# item category characteristic curves
with(data, model_grm_plot(a, b, type='prob', total=FALSE))
# item category information funciton curves
with(data, model_grm_plot(a, b, type='info', total=FALSE))
```


## Estimation 

IRT models describe the interaction between people and items using probabilistic functions. However, neither people nor item parameters are naturally known in reality and they have to be estimated based on the observed response data. Thus, parameter estimation, which calibrates items and scores response vectors, is the centerpiece of data analysis in IRT.

Here are some examples of estimating model parameters using the 3PL model in this package.

```{r, fig.width=10, fig.height=3}
# generate data
data <- model_3pl_gendata(2000, 40)
# score response vectors using EAP
scores <- with(data, model_3pl_eap(u, a, b, c))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
# score response vectors using MAP
scores <- with(data, model_3pl_map(u, a, b, c))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
# calibrate items using joint MLE
calib <- with(data, model_3pl_jmle(u, true_params=data))
# calibrate items using marginal MLE
calib <- with(data, model_3pl_mmle(u, true_params=data))
# calibrate items using joint MLE and fix c=0
calib <- with(data, model_3pl_jmle(u, c=0, true_params=data))
# calibrate items using joint MLE and fix a=1 and c=0
calib <- with(data, model_3pl_jmle(u, a=1, c=0, true_params=data))
```

Here are examples of estimating parameters using the GPCM model.

```{r, fig.width=10, fig.height=3}
# generate data
data <- model_gpcm_gendata(1000, 20, n_c=3)
# score response vectors using EAP
scores <- with(data, model_gpcm_eap(u, a, b, d))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
# score response vectors using MAP
scores <- with(data, model_gpcm_map(u, a, b, d))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
# calibrate items using joint MLE
calib <- with(data, model_gpcm_jmle(u, true_params=data))
# calibrate items using marginal MLE
calib <- with(data, model_gpcm_mmle(u, true_params=data))
```

Here are examples of estimating parameters using the GRM model.

```{r, fig.width=10, fig.height=3}
# generate data
data <- model_grm_gendata(1000, 20, n_c=3)
# score response vectors using EAP
scores <- with(data, model_grm_eap(u, a, b))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
# score response vectors using MAP
scores <- with(data, model_grm_map(u, a, b))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
# calibrate items using joint MLE
calib <- with(data, model_grm_jmle(u, true_params=data))
# calibrate items using marginal MLE
calib <- with(data, model_grm_mmle(u, true_params=data))
```

Here are examples of estimating parameters using the mixed model.

```{r}
# generate data
data <- model_mixed_gendata(2000, n_3pl=40, n_gpcm=10, n_grm=10, n_c=3)
# score response vectors using EAP
scores <- with(data, model_mixed_eap(u, items))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
# score response vectors using MAP
scores <- with(data, model_mixed_map(u, items))
round(c(corr=cor(data$t, scores$t), rmse=rmse(data$t, scores$t)), 2)
```


## Getting help

If you encounter a bug, please post a code example that exposes the bug on [github](https://github.com/xluo11/Rirt/issues). You can post your questions and feature requests on [github](https://github.com/xluo11/Rirt/issues) or contact the [author](mailto:xluo1986@gmail.com).